Open archaeology, open source?

### 2024-05-23 analysis/_06-network.qmd
Explain regression line
Revisions based on editorial comments.

By contributing to shared repositories—whether with code (commits), issues, or comments—archaeologists using GitHub form a collaborative network which we can map using data from the GitHub API.
Here we consider two facets of this network: repositories connected by common contributors (the repository–repository graph), and users connected by contributions to common repositories (the user–user graph).
In both cases, number of contributions constitutes a natural measure of the strength or weight of the connection, which can be further broken down by type of contribution (commit, issue/pull request, or comment).

@@

Our data shows that there is a significant network of archaeologists collaborating on GitHub.
`r percent(graph_n_repos / graph_n_repos_total)` of repositories and `r percent(graph_n_users / graph_n_users_total)` of users in our dataset are connected to at least one other repository or user.
Of these, `r percent(max(subgraphs_n_repos$n) / graph_n_repos)` of repositories and `r percent(max(subgraphs_n_users$n) / graph_n_users)` of users belong to a single connected subgraph (@fig-graph-repo-user and @fig-graph-repo-repo).

We delimited `r nrow(filter(clusters_n_repos, n > 1))` distinct clusters that outline the topography of the repository–repository network.
While many of these clusters are interconnected, some discrete components containing between 2-20 repositories appear as distinct from a primary core.
The core cluster is characterized by repositories whose contributors commit to projects other than their own, and it includes a smorgasbord of projects whose contributors share varied interests.

Clustering also reveals distinct collaborative networks within the user-user graph.
We again see a complementary primary core connected to several more peripheral clusters, which are internally-cohesive and exhibit few connections with other peripheral clusters.
The central core bridges all the peripheral clusters.
The central core is not uniform, and comprises several relatively discrete clusters representing collaborative sub-communities.
While these clusters are internally cohesive, they exhibit enough connections to other members of the central core so as to not be considered as separate or peripheral clusters.

@@

```{r fig-repo-betweenness-age-length}
#| fig-cap: Repository centrality by age (left) and length (right). Centrality is measured by node betweenness weighted by number of contributions.
left <- oarch_graph_repo_repo |>
  as_tibble() |> 
  left_join(oarch, by = c("repo" = "gh_repo")) |>
  mutate(
    age = map_int(gh_commits, \(x) min(today() - date(x$datetime)))
  ) |>
  ggplot(aes(age, betweenness)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10(labels = NULL) +
  geom_smooth(method = "glm") +
  labs(x = "Days", y = "Log centrality →") +
  theme_oarch()

@@

When comparing across categories and platforms, the highest mean centrality is seen in repositories that contain lists and datasets, standalone software, or packages and libraries, and in repositories based on Python, R or QGIS (@fig-repo-betweenness-category-platform).
Interestingly, these trends depart from the observed popularity of different categories and platforms in the open-archaeo dataset as a whole (see @sec-open-archaeology): standalone software is more central than packages/libraries, even though there are more of the latter by a significant margin.
This may be due to the fact that many packages are developed to support specific practices or use-cases (often inspired by personal need), or are designed to run relatively stable statistical functions that need not change over time.
These are therefore relatively stable and require little additional input after release.
On the other hand, as discussed in @sec-open-archaeology, standalone software tend to integrate multiple system components and may evolve over time to add new features or support new workflows.
Moreover, standalone software are generally rooted in longer-term and community-held objectives, and their development may therefore be backed by insitutions with funding and resources to support developers.

Despite being a minority language, Blender packages are more central than all other package platforms on average, but this is a statistical anomaly caused by uneven sampling.
R is naturally the platform with the next highest average centrality since it serves as a _lingua franca_ that draws developers from across the discipline.
Many of the QGIS plugins add various specialized features to the extensible GIS platform, and are therefore developed by interdisciplinary teams, which explains its high rank.
Python projects, which tend to be infrastructural or are of interest to members of other fields, are also highly ranked in terms of average centrality.
Our findings support the notion that there are significant development patterns when working across different languages and platforms, and further analysis to qualify these observation is warranted.

```{r fig-user-betweenness-contribs}
#| fig-cap: User centrality by total number of contributions. Centrality is measured by node betweenness weighted by number of contributions.
oarch_graph_user_user |>
  as_tibble() |> 
  left_join(oarch_contributors, by = c("user" = "contributor")) |>
  ggplot(aes(contribs, betweenness)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "glm") +
  labs(x = "Total contributions", y = "Centrality") +
  theme_oarch()
```

>>>>

By contributing to shared repositories—whether with code (commits), issues, or comments—archaeologists using GitHub form a collaborative network which we can map using data from the GitHub API.
From a two-mode network with edges representing users’ contributions to repositories (Figure 6), we derived two one-mode networks that render co-contributions as connections in themselves: repositories connected by common contributors (the repository–repository graph, presented in Figure 7), and users connected by contributions to common repositories (the user–user graph, presented in Figure 8).
In both cases, number of contributions constitutes a natural measure of the strength or weight of the connection, which can be further broken down by type of contribution (commit, issue/pull request, or comment).
We identified clusters using the edge-betweenness community detection method, which operates by locating the edges that are situated along the shortest paths between all pairs of nodes, and which therefore exhibit high betweenness centrality; these edges form indirect links between otherwise completely unconnected sections of the network, and as such are the loci that distinguish groups of nodes that are more highly connected to each other than they are to others [@girvan2002].

@@

Our data show that there is a significant network of archaeologists collaborating on GitHub.
`r percent(graph_n_repos / graph_n_repos_total)` of repositories and `r percent(graph_n_users / graph_n_users_total)` of users in our dataset are connected to at least one other repository or user.
Of these, `r percent(max(subgraphs_n_repos$n) / graph_n_repos)` of repositories and `r percent(max(subgraphs_n_users$n) / graph_n_users)` of users belong to a single connected subgraph (@fig-graph-repo-user and @fig-graph-repo-repo).
This indicates that most repositories that have had more than one contributor, or whose contributors have worked independently on more than one repository, are not isolated, and that at least one of their contributors have, at least one time, worked with members of a broader interconnected population.

We delimited `r nrow(filter(clusters_n_repos, n > 1))` distinct clusters that outline the topography of the repository–repository network (@fig-graph-repo-repo).
While many of these clusters are interconnected, some discrete components containing between 2-20 repositories appear as distinct from a primary core.
The core cluster is characterized by repositories whose contributors commit to projects other than their own, and it includes a smorgasbord of projects whose contributors share varied interests.

Clustering also reveals distinct collaborative networks within the user-user graph (@fig-graph-user-user).
We again see a complementary primary interconnected subgraph and several unconnected subgraphs.
The primary subgraph comprises several connected clusters, including one central cluster and several more peripheral clusters, which are internally-cohesive and exhibit few connections with other peripheral clusters.
The central cluster bridges all the peripheral clusters.
Moreover, the central cluster is not uniform, and comprises several relatively discrete components representing collaborative sub-communities.
While these components are internally cohesive, they exhibit enough connections to other members of the central cluster so as to not be considered as separate or peripheral clusters.

@@

```{r fig-repo-betweenness-age-length}
#| fig-cap: Repository centrality by age (left) and length (right). Centrality is measured by node betweenness weighted by number of contributions. Blue line indicates a generalized linear regression, with confidence envelope shaded in grey.
left <- oarch_graph_repo_repo |>
  as_tibble() |> 
  left_join(oarch, by = c("repo" = "gh_repo")) |>
  mutate(
    age = map_int(gh_commits, \(x) min(today() - date(x$datetime)))
  ) |>
  ggplot(aes(age, betweenness)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10(labels = NULL) +
  geom_smooth(method = "glm") +
  labs(x = "Days", y = "Log centrality →") +
  theme_oarch()

@@

When comparing across categories and platforms, the highest mean centrality is seen in repositories that contain lists and datasets, standalone software, or packages and libraries, and in repositories based on Python, R or QGIS (@fig-repo-betweenness-category-platform).
Interestingly, these trends depart from the observed popularity of different categories and platforms in the open-archaeo dataset as a whole (see @sec-open-archaeology): standalone software is more central than packages/libraries, even though there are more of the latter by a significant margin.
This may be due to the fact that many packages are developed to support specific practices or use-cases (often inspired by personal need), or are designed to run relatively stable statistical functions that need not change over time.
These are therefore relatively stable and require little additional input after release.
On the other hand, standalone software tend to integrate multiple system components and may evolve over time to add new features or support new workflows.
Moreover, standalone software are generally rooted in longer-term and community-held objectives, and their development may therefore be backed by insitutions with funding and resources to support developers.

Despite being a minority language, Blender packages are more central than all other package platforms on average, but this is a statistical anomaly caused by uneven sampling.
R is naturally the platform with the next highest average centrality since it serves as a _lingua franca_ that draws developers from across the discipline.
Many of the QGIS plugins add various specialized features to the extensible GIS platform, and are therefore developed by interdisciplinary teams, which explains its high rank.
Python projects, which tend to relate to the development of information infrastructures or are of interest to members of other fields (e.g., palaeo-ecology, other fieldwork-based disciplines), are also highly ranked in terms of average centrality.
Further analysis is warranted to qualify these observations regarding the significance of development patterns when working across different languages and platforms.

```{r fig-user-betweenness-contribs}
#| fig-cap: User centrality by total number of contributions. Centrality is measured by node betweenness weighted by number of contributions.  Blue line indicates a generalized linear regression, with confidence envelope shaded in grey.
oarch_graph_user_user |>
  as_tibble() |> 
  left_join(oarch_contributors, by = c("user" = "contributor")) |>
  ggplot(aes(contribs, betweenness)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "glm") +
  labs(x = "Total contributions", y = "Centrality") +
  theme_oarch()
```

### 2024-05-23 analysis/_07-conclusion.qmd
Revisions based on editorial comments.

Our goal in this study was to investigate the under-explored research practices involved in research software engineering in archaeology.
We sought to identify not only _what_ kinds of software archaeologists are making, but _how_ archaeologists create these tools as part of a broader community of practice. 
Our emphasis on the collaborative experiences involved in open source software development emerged from our experience maintaining open-archaeo, through which we observed that making one's code openly available on the web does not necessarily garner the benefits often touted by open science advocates, namely that source code can be audited, forked, and appropriated for alternative use cases, which are effectively social and collaborative experiences.

@@

There is an emerging community of practice around open source research software in archaeology. 
All but a handful of the GitHub repositories we analysed have more than commit, showing that archaeologists use it for ongoing work rather than merely to upload finished products.
They relatively frequently make use of the 'star' and 'comment' features to engage with others' repositories (@fig-collab-features) and, via these and other shared contributions, we can trace a collaborative network that includes the majority of archaeologists active on GitHub (see @sec-network).

@@

Our network analysis (@sec-network) similarly draws attention to the real-world collaborative ties that underpin archaeological open source software development.
We identify a core cluster representing a series of collaborative ties among members of an archaeological software engineering community of practice.
This core exhibits complexity that corresponds with social patterns, such as the presence of various clusters representing interconnected interest or affinity groups.
Indeed, we have found that 'real-world' social connections and institutional support structures are strong predictors of centrality, since these clusters are representative of established professional partnerships.
This suggests that archaeological open source is firmly embedded within existing power structures that permeate academic life, both online and offline.
Similarly, we found that the individuals who play critical roles in supporting the archaeological open source community are precariously employed workers. 
Far from open source being inherently distributed, resilient, and open-ended, this indicates that research software engineering is actually quite centralized, fragile, and based heavily on existing professional connections and endeavours.

These findings call into question the notion that archaeologists benefit from the positive outcomes that are commonly argued to be the natural results of open source development models -- namely, greater degrees of extensibility and participatory action.
While opening the source code may facilitate these positive outcomes as necessary preconditional factors, we argue that this only amounts to establishing the _potential_ for people to put these values into practice.
We argue that the objectives and circumstances that frame archaeological practice significantly influence how far archaeologists (and academics in general) are willing to push for these values, and limit the ability for archaeologists to do open source in ways that resemble more mainstream open source projects.
For instance, successful open source projects like the Linux kernel, openSSL, or the Firefox web browser are driven by collective and popular interest in ensuring that code remains functional, and the code base is therefore constantly in flux and bears an accumulating list of contributing members.
This differs from the organizational principles that govern much archaeological work, namely where a director or directors (of a field project, research group, etc.) sets the goals and orientation of the group and commissions and manages other actors accordingly. 
Moreover, archaeological projects ultimately seek to produce stable textual outcomes bearing clear delineation of authorship and that require no upkeep whatsoever. 
Sustaining an open source project is simply not compatible with the factors that currently drive the momentum behind archaeological work.

As such, we advocate for more focused attention on specific disciplinary norms and institutional support structures that inform how knowledge is created and validated, and how varied contributions to the scholarly enterprise are mediated, credited, and valued.
In other words, if we want to make open source effective in relation to the aforementioned goals of encouraging greater inclusivity, transparency, and productivity, we also need to foster a culture that supports active, pragmatic and humble critique, and which instills a de-territorialized attitude concerning what it means to contribute to collective knowledge [@morgan2015; @ducke2015].
This means fighting against the pathological power-relations that scaffold all aspects of academic life, and not fooling ourselves into believing that technical solutions (i.e. using git) will, on their own, resolve the wicked social problems that lie at the heart of scientific research practice.

>>>>

Our goal in this study was to investigate the under-explored research practices involved in research software engineering in archaeology.
We sought to identify not only _what_ kinds of software archaeologists are making, but _how_ archaeologists create these tools as part of a broader community of practice. 
Our emphasis on the collaborative experiences involved in open source software development emerged from our experience maintaining open-archaeo.
We observed that making one's code openly available on the web does not necessarily garner the benefits often touted by open science advocates, namely that source code can be audited, forked, and appropriated for alternative use cases, which are effectively social and collaborative experiences.

@@

There is an emerging community of practice around open source research software in archaeology. 
All but a handful of the GitHub repositories we analysed have more than one commit, showing that archaeologists use it for ongoing work rather than merely to upload finished products.
They relatively frequently make use of the 'star' and 'comment' features to engage with others' repositories (@fig-collab-features) and, via these and other shared contributions, we can trace a collaborative network that includes the majority of archaeologists active on GitHub (see @sec-network).

@@

Our network analysis (@sec-network) similarly draws attention to the real-world collaborative ties that underpin archaeological open source software development.
We identify a core cluster representing a series of collaborative ties among members of an archaeological software engineering community of practice.
This core exhibits complexity that corresponds with social patterns, such as the presence of various clusters representing interconnected interest or affinity groups.
Indeed, we have inferred that 'real-world' social connections and institutional support structures are strong predictors of centrality, since these clusters are representative of established professional partnerships.
This suggests that archaeological open source is firmly embedded within existing power structures that permeate academic life, both online and offline.
Similarly, we found that the individuals who play critical roles in supporting the archaeological open source community are precariously employed workers. 
Far from open source being inherently distributed, resilient, and open-ended, this indicates that research software engineering is actually quite centralized, fragile, and based heavily on existing professional connections and endeavours.

These findings call into question the notion that archaeologists benefit from the positive outcomes that are commonly argued to be the natural results of open source development models -- namely, greater degrees of extensibility and participatory action.
While opening the source code may facilitate these positive outcomes as necessary preconditional factors, we argue that this only amounts to establishing the _potential_ for people to put these values into practice.
We argue that the objectives and circumstances that frame archaeological practice significantly influence how far archaeologists (and academics in general) are willing to push for these values, and limit the ability for archaeologists to do open source in ways that resemble more mainstream open source projects.
For instance, successful open source projects like the Linux kernel, openSSL, or the Firefox web browser are driven by collective and popular interest in ensuring that code remains functional, and the code base is therefore constantly in flux and bears an accumulating list of contributing members.
This differs from the organizational principles that govern much archaeological work, namely where a director or directors (of a field project, research group, etc.) sets the goals and orientation of the group and commissions and manages other actors accordingly. 
Moreover, archaeological projects ultimately seek to produce stable textual outcomes bearing clear delineation of authorship and that require no upkeep whatsoever. 
Sustaining an open source project is simply not compatible with the factors that currently drive the momentum behind archaeological work.
This is compounded by the fact that many archaeologists consider software development as a form of support work and of lesser value than traditional academic research activities, and that research software engineers are often precariously employed [@baxter2012; @cornelius-bell2021].

As such, we advocate for more focused attention on specific disciplinary norms and institutional support structures that inform how knowledge is created and validated, and how varied contributions to the scholarly enterprise are mediated, credited, and valued [@leonelli2023; @bennett2021; @khan2024].
In other words, if we want to make open source effective in relation to the aforementioned goals of encouraging greater inclusivity, transparency, and productivity, we also need to foster a culture that supports active, pragmatic and humble critique, and which instills a de-territorialized attitude concerning what it means to contribute to collective knowledge [@morgan2015; @ducke2015].
This means fighting against the pathological power-relations that scaffold all aspects of academic life, and not fooling ourselves into believing that technical solutions (i.e. using git) will, on their own, resolve the wicked social problems that lie at the heart of scientific research practice.
